Inferring the existence and direction of causal associations in the face of measurement error

### 2016-12-31 docs/manuscript.tex
updated

\setlength{\droptitle}{-2em}
  \title{Inferring the existence and direction of causal associations in the face
of measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{29 December 2016}

@@

The last decade has seem a rapid acceleration in technological advances
that enable us to assess the human phenome in ever increasing detail,
furnishing biologists with the tantalising opportunity to map the
molecular basis of complex traits. This does, however, require
overcoming the statistical challenge of correctly defining the causal
architectures that underlie these high dimensional data. For example,
identifying metabolic perturbations that lead to
disease\textsuperscript{1}, or understanding the influence of epigenomic
influences on gene expression\textsuperscript{2} all depend on
understanding the causal nature of any putative observational
associations.

@@

Observational study data are now widely available but notoriously
susceptible to pitfalls in making causal inference, with basic
regression-based techniques unable to distinguish a true causal
association from reverse causation or confounding\textsuperscript{3--5}.
In response to this, genetic instrumentation has emerged as a valuable
technique for improving the reliability of causal inference in
observational data, and with the coincident rise in genome-wide
association studies it is now a prominent tool that is applied in
several different guises, and has been applied many forms of high
throughput 'omic variables. However, an often neglected pitfall to these
approaches is the influence of non-differential measurement error on the
reliability of causal inference.

Non-differential measurement error (or observational error, henceforth
referred to simply as measurement error) is the difference between the
measured value of a quantity and its true value. Such variability can
arise through a whole plethora of mechanisms, which are often unique to
the study design and difficult to avoid\textsuperscript{6,7}. Array
technology is now commonly used to obtain high throughput phenotyping at
low cost, but they come with the problem of having imperfect resolution,
for instance methylation levels as measured by the Illumina450k chip are
prone to have some amount of noise around the true
value\textsuperscript{8,9}. Relatedly, if the measurement of biological
interest is the methylation level in a T cell, then measurement error of
this value can be introduced by using methylation levels from whole
blood samples because the measured value will be an assay of many cell
types\textsuperscript{10}.

Measurement error will of course arise in more low-tech data too, for
example when measuring BMI one is typically interested in using this as
a proxy for adiposity, but it is clear that the correlation between BMI
and underlying adiposity is not perfect\textsuperscript{11}. A similar
problem of biological misspecification is unavoidable in disease
diagnosis, and measuring behaviour such as smoking or diet is
notoriously difficult to do accurately. Measurement error can also be
introduced after the data has been collected, for example the
transformation of non-normal data for the purpose of statistical
analysis will lead to a new variable that will typically incur both
changes in scale and imprecision (noise) compared to the original
variable. The sources of measurement error are not limited to this
list\textsuperscript{7}, and its impact has been explored in the
epidemiological literature extensively\textsuperscript{12--14}.

@@

Mediation-based approaches employ genetic instruments (typically single
nucleotide polymorphisms, SNPs) to orient the causal direction between
the exposure and the outcome. If a SNP is associated with an exposure,
and the exposure is associated with some outcome, then it logically
follows that the indirect influence of the SNP on the outcome will
disappear when conditioning on the exposure. Here, the exposure
`mediates' the association between the SNP and the outcome, providing
information about the causal influence of the exposure on the outcome.
This forms the basis of a number of methods such as genetical
genomics\textsuperscript{15}, the regression-based causal inference test
(CIT)\textsuperscript{16,17}, a structural equation modelling (SEM)
implementation in the NEO software\textsuperscript{18}, and various
other methods including Bayesian approaches\textsuperscript{19}. They
have been employed by a number of recent publications that make causal
inferences in large scale `omics datasets\textsuperscript{19--24}.

@@

Such is not the case for MR, where in order to infer the direction of
causality the instrument must be known to influence the exposure
directly, associating with the outcome only through the exposure. This
requirement can be problematic because if there exists a putative
association between two variables, with the instrumental SNP being
robustly associated with each, it can be difficult to determine which of
the two variables is subject to the direct effect of the SNP (i.e., for
which of the two variables is the SNP a valid instrument? Figure 1). By
definition, we expect that if the association is causal then a SNP for
the exposure will be associated with the outcome, so if the researcher
erroneously uses the SNP as an instrument for the outcome then they are
likely to see an apparently robust causal association of outcome on
exposure. Such a situation can arise in many scenarios. Genome-wide
association studies (GWASs) that identify genetic associations for
complex traits are, by design, hypothesis free and agnostic of genomic
function, and it often takes years of follow up studies to understand
the biological nature of a putative GWAS hit\textsuperscript{25}. For
example, the same FTO variant that is associated with body mass index
(BMI) is also associated with C-reactive protein (CRP). It is known that
this is because BMI causes CRP\textsuperscript{26}, but supposing we
only had \emph{a priori} knowledge that FTO associated with CRP, and we
used it to instrument CRP in an MR analysis of CRP against BMI, we would
mistakenly infer that CRP causes BMI. Another situation is where the
causal directions between 'omic measures need to be determined, for
example if a DNA methylation probe is associated with expression of an
adjacent gene, then is a cis-acting SNP an instrument for the DNA
methylation level, or the gene expression level\textsuperscript{27}
(Figure 1)?

MR does however have some important advantages over mediation based
approaches. First, MR is used as a tool to not just deduce the existance
of causality, but to also estimate the causal effect size. Second, using
mediation requires that the exposure, outcome and instrument variables
are all measured in the same data, whereas recent extensions to MR
circumvent this requirement, allowing causal inference to be drawn when
exposure variables and outcome variables are measured in different
samples\textsuperscript{28}. This has the crucial advantage of improving
statistical power by allowing analysis in much larger sample sizes, and
dramatically expands the breadth of possible phenotypic relationships
that can be evaluated. Third, MR is substantially more robust to there
being measurement error in the exposure and outcome variables. Indeed
instrumental variable (IV) analysis was in part initially introduced as
a correction for measurement error\textsuperscript{29}, whereas it has
been noted that both classic mediation-based
analyses\textsuperscript{13,14,30,31} and mediation-based methods that
use instrumental variables\textsuperscript{32,33} are prone to be
unreliable in its presence. We argue that this is a serious limitation.

Here we show using theory and simulations how non-differential
measurement error and measurement imprecision can lead to unreliable
causal inference in the mediation-based CIT method. We then present an
extension to MR that allows researchers to ascertain the causal
direction of an association even when the biology of the instruments are
not fully understood, and to evaluate the sensitivity of the result to
measurement error. This improves the utility of MR in cases where
mediation based methods might have otherwise been used preferentially.
Finally, we apply this new method to infer the direction of causation
between DNA methylation levels and gene expression levels.

@@

The analysis proceeds by simulating variables that have known causal
effects and valid instruments, and then applying measurement error in
different configurations. We then assess the performance of the CIT test
and MR analysis for a) inference of a causal relationship and b)
inferring the correct direction of causality. Included in the MR
analysis is a simple extension that is used to make inference of the
causal direction when it is unknown which of the correlated variables
the instrument is directly acting upon. All analysis was performed using
the R programming language\textsuperscript{34} and code is made
available at \href{}{github url to go here}.

@@

The CIT method\textsuperscript{16} is implemented in the R package
\emph{R/cit}\textsuperscript{17}. The methodology of the CIT is as
follows. Assume an exposure \(x\) is instrumented by a SNP \(g\), and
the exposure \(x\) is causally related to an outcome \(y\). Thus,

@@

where \(\hat{x} = \hat{\alpha}_g + \hat{\beta}_g g\), a causal
relationship is inferred if the null hypothesis that \(\beta_{MR} = 0\)
is rejected. A p-value for this test, \(p_{MR}\), was obtained using the
R package \(R/systemfit\)\textsuperscript{35}.

The direction of the causal association was inferred as follows.
Assuming no horizontal pleiotropy (i.e. \(g\) only has an effect on one
of the variables through the other variable) it is desirable to know
which of the variables, \(x\) or \(y\), it has a direct influence on.
This can be achieved by assessing which of the two variables has the
biggest absolute correlation with \(g\) (Appendix 2). This test can be
formalised by testing for a difference in the correlations \(\rho_{gx}\)
and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
within a population\textsuperscript{36}. It is calculated as

@@

Note that the same correlation test approach can be applied to a
two-sample MR\textsuperscript{28} setting. Two-sample MR refers to the
case where the SNP-exposure association and SNP-outcome association are
calculated in different samples (or from publicly available summary
statistics). Here the Steiger test of two independent correlations can
be applied where.

@@

An advantage of using the Steiger test in the two sample context is that
it can compare correlations in independent samples where sample sizes
are different. Steiger test statistics were calculated using the
\emph{r.test} function in the R package
\emph{R/psych}\textsuperscript{37}.

@@

Two sample MR\textsuperscript{28} was performed using the summary
statistics presented in\textsuperscript{27} for each of 4122
associations between DNA methylation and gene expression levels. Because
only summary statistics were available (effect, standard error, effect
allele, sample size, p-values) for the instrumental SNP on the
methylation and gene expression levels, the Steiger test of two
independent correlations was used to infer the direction of causality
for each of the associations. The Wald ratio test was then used to
estimate the causal effect size for the estimated direction for each
association.

@@

This is a simple method to orient the direction of causality in an MR
analysis when the underlying biology of the SNP is not fully understood.
But it can be shown that in the presence of measurement imprecision,
\(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\) (Appendix 2), where \(d\)
is a metric of the degree to which the Steiger test is liable to provide
the wrong estimate (\emph{i.e.} if \(d>0\) then the Steiger test is
likely to be correct about the causal direction). Thus it is important
to note that under certain conditions when measurement imprecision is
present the Steiger test could make erroneous inference about the causal
direction because the value of \(d\) is not restricted to being greater
than 0. However Figure 7 shows that when there is no measurement error
in \(y\), the Steiger test is unlikely to infer the wrong direction of
causality even if there is measurement error in \(x\). It also shows
that in most cases where \(y\) is measured with error, especially when
the causal effect between \(x\) and \(y\) is not very large, the
condition of \(d > 0\) is satisfied.

@@

We used the Steiger test to infer the direction of causality between DNA
methylation levels and gene expression levels. To do this we obtained a
list of 458 gene expression - DNA methylation associations as reported
in Shakhbazov et al\textsuperscript{27}. These were filtered to be
located on the same chromosome, have robost correlations after
correcting for multiple testing, and to share a SNP that had a robust
cis-acting effect on both the DNA methylation probe and the gene
expression probe. Using the reported summary data we used the Steiger
test of two independent correlations to identify which direction of
causality was most likely for each association. We found that the causal
direction commonly goes in both directions (Table 1) but methylation
levels were more likely to influence expression levels than vice versa
(probability of expression causing methylation = 0.0163873 (95\% CI
0.3738727 - 0.4874269) amongst NA associations with
\(p_{Steiger} < 0.05\)).

We performed two sample MR\textsuperscript{28} for each association in
the direction of causality that we estimated to be the correct direction
from the Steiger test. We observed that the sign of the MR estimate was
always in the same direction as the Pearson correlation coefficient
reported by Shakhbazov et al\textsuperscript{27}, and that there was a
moderate correlation between the magnitude of the causal correlation and
the observational Pearson correlation (Table 1).

We also observed that for associations where methylation caused gene
expression the causal effect was more likely to be negative than for the
associations where gene expression caused methylation (OR = 0.5326663
(95\% CI 0.3172203 - 0.8851868)), suggesting that reducing gene
expression levels at a controlling CpG typically leads to increased gene
expression levels, consistent with expectation\textsuperscript{2}.

@@

In the epidemiological literature issues of measurement error in
mediation is relatively well explored\textsuperscript{12}. Our analysis
extends this to related methods such as CIT that are widely used in
predominantly 'omic data. These methods are indeed liable to the same
issues as standard mediation based analysis, and specifically we show
that as measurement error in the exposure variable increases, CIT is
likely have reduced statistical power, and liable to infer the wrong
direction of causality. We also demonstrate that, though seemingly
unintuitive, increasing sample size does not resolve the issue, rather
it leads to more extreme p-values for the model that predicts the wrong
direction of causality.

@@

Mediation based network approaches are very well
established\textsuperscript{33} and have a number of extensions that
make them valuable tools, including for example network construction.
But because they are predicated on the basic underlying principles of
mediation they are liable to suffer from the same issues of measurement
error. Recent advances in MR methodology, for example applying MR to
genetical genomics\textsuperscript{38}, multivariate
MR\textsuperscript{39}, network MR\textsuperscript{40} and mediation
through MR\textsuperscript{41} may offer more robust alternatives for
these more complicated problems.

In our simulations we focused on the simple case of a single instrument
in a single sample setting, and assumed that pleiotropy (the influence
of the instrument on the outcome through a mechanism other than the
exposure) was not present. Recent method developments in
MR\textsuperscript{42} have focused on accounting for the issues that
horizontal pleiotropy can introduce, but how they perform in the
presence of measurement error remains to be explored. An important
advantage that MR confers over most mediation based analysis is that it
can be performed in two samples, which can considerably improve power
and expand the scope of analysis. However, whether there is a
substantive difference in two sample MR versus one sample MR in how
measurement error has an effect is not yet fully understood. We have
also assumed no measurement error in the genetic instrument, which is
not unreasonable given the strict QC protocols that ensure high quality
genotype data is available to most studies. We have restricted the scope
to only exploring non-differential measurement error and avoided the
complications incurred if measurement error in the exposure and outcome
is correlated, however our analysis goes beyond many previous
explorations of measurement error by assessing the impacts of both
imprecision (noise) and linear transformations of the true variable on
causal inference.

@@

\includegraphics{manuscript_files/figure-latex/steiger_sensitivity_plot-1.pdf}

Figure 3: a) We can predict the values the Steiger test would take
(z-axis) for different potential values of measurement error (x and y
axes), drawn here as the blue surface. When \(\rho_{g,y} > \rho_{g,x}\),
as denoted by range of values where the blue surface is above the red
plane, those values of measurement error lead to our observed Steiger
test inferring the wrong causal direction. Where the blue surface lies
below the red plane, these measurement error values support the inferred
causal direction of X to Y. A measure of sensitivity, therefore, is the
volume of the positive space bound by the blue surface as a proportion
of the total space bound by the blue surface. Sensitivity of 0 denotes
no measurement error values that could lead to inferrence of incorrect
direction, whereas the maximum value of 0.5 denotes measurement error
equal chance of either direction across all measurement error ranges. In
this case, where \(\rho_{g,x}^2 = 0.01\) and \(\rho_{x,y}^2 = 0.1\), the
sensitivity is 0.18. b) The influence of \(\rho_{x,y}\) (x-axis) on the
Steiger sensitivity values (y-axis) for different values of
\(\rho_{g,x}^2\). Inferring the direction of causality is more
susceptible to measurement error when there are stronger causal
associations between X and Y.

@@

In the presence of measurement error in \(x\) and \(y\), however, the
Steiger test will instead be assessing the inequality
\(\rho_{g, x_o} > \rho_{g, y_o}\). In order to assess how sensitive the
Steiger test is to measurement error, we can evaluate the range of
potential values of measurement error in \(x\) and \(y\) over which the
empirical result from the Steiger test would return the wrong value.

@@

Where \(z = \rho_{g,y} - \rho_{g,x} < 0\) the measurement error range in
\(x_o\) and \(y_o\) will not change the qualitative inference of the
direction of causality. To evaluate the sensitivity, \(S\), of the
Steiger test to measurement error, the objective is to calculate the
proportion of the total absolute volume of this function that is above
the \(z = 0\) plane,

\[
S = \frac{V_{z>0}}{2V_{z>0} + |V_{z}|}
\]

\(V_{z}\), the total volume of the function (Figure 3), can be obtained
analytically by solving:

@@

\(V_{z>0}\), the proportion of the volume that lies above the \(z=0\)
plane, can also be obtained analytically. The region of this volume is
bound by the values of \(\rho_{x,x_o}\) and \(\rho_{y,y_o}\) where
\(0 = \rho_{g,y} - \rho_{g,x}\), which can be expanded to
\(\rho_{y,y_o} = \rho_{g,y_o}\rho_{x,x_o} / \rho_{g,x_o}\). Hence,

\[
\begin{aligned}
V_{z>0} & = \int^1_{\rho_{g,x_o}} \int^{\frac{\rho_{g,y_o}\rho_{x,x_o}}{\rho_{g,x_o}}}_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
& = 2\rho_{g,x_o}\rho_{g,y_o} - 2\rho_{g,y_o} - \rho_{g,y_o}log(\rho_{g,x_o}) - \rho_{g,x_o}\rho_{g,y_o}log(\rho_{g,x_o})
\end{aligned}
\]

The sensitivity, \(S\), can take values from 0 to 0.5, where low values
suggest there are

@@

\hypertarget{refs}{}
\hypertarget{ref-Kettunen2012}{}
1. Kettunen, J. \emph{et al.} Genome-wide association study identifies
multiple loci influencing human serum metabolite levels. \emph{Nature
genetics} \textbf{44,} 269--76 (2012).

>>>>

\setlength{\droptitle}{-2em}
  \title{Inferring the existence and direction of causal associations in the face
of measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{31 December 2016}

@@

@@

Observational measures of the human phenome are growing ever more
abundant, but using these data to make causal inference is notoriously
susceptible to many pitfalls, with basic regression-based techniques
unable to distinguish a true causal association from reverse causation
or confounding\textsuperscript{1--3}. In response to this, genetic
instrumentation has emerged as a technique for improving the reliability
of causal inference in observational data, and with the coincident rise
in genome-wide association studies it is now a prominent tool that is
applied in several different guises\textsuperscript{3--6}. However,
potential pitfalls still persist and one that is often neglected is the
influence of non-differential measurement error on the reliability of
causal inference.

Non-differential measurement error (or observational error, henceforth
referred to simply as measurement error) is the difference between the
measured value of a quantity and its true value. Such variability can
arise through a whole plethora of mechanisms, which are often unique to
the study design and difficult to avoid\textsuperscript{7,8}. Array
technology is now commonly used to obtain high throughput phenotyping at
low cost, but they come with the problem of having imperfect resolution,
for instance methylation levels as measured by the Illumina450k chip are
prone to have some amount of noise around the true
value\textsuperscript{9,10}. Relatedly, if the measurement of biological
interest is the methylation level in a T cell, then measurement error of
this value can be introduced by using methylation levels from whole
blood samples because the measured value will be an assay of many cell
types\textsuperscript{11}.

Measurement error will of course arise in more low-tech data too, for
example when measuring BMI one is typically interested in using this as
a proxy for adiposity, but it is clear that the correlation between BMI
and underlying adiposity is not perfect\textsuperscript{12}. A similar
problem of biological misspecification is unavoidable in disease
diagnosis, and measuring behaviour such as smoking or diet is
notoriously difficult to do accurately. Measurement error can also be
introduced after the data has been collected, for example the
transformation of non-normal data for the purpose of statistical
analysis will lead to a new variable that will typically incur both
changes in scale and imprecision (noise) compared to the original
variable. The sources of measurement error are not limited to this
list\textsuperscript{8}, and its impact has been explored in the
epidemiological literature extensively\textsuperscript{13--15}.

@@

Mediation-based approaches employ genetic instruments (typically single
nucleotide polymorphisms, SNPs) to orient the causal direction between
the exposure and the outcome. If a SNP is associated with an exposure,
and the exposure is associated with some outcome, then it logically
follows that the indirect influence of the SNP on the outcome will
disappear when conditioning on the exposure. Here, the exposure
`mediates' the association between the SNP and the outcome, providing
information about the causal influence of the exposure on the outcome.
This forms the basis of a number of methods such as genetical
genomics\textsuperscript{16}, the regression-based causal inference test
(CIT)\textsuperscript{4,17}, a structural equation modelling (SEM)
implementation in the NEO software\textsuperscript{5}, and various other
methods including Bayesian approaches\textsuperscript{6}. They have been
employed by a number of recent publications that make causal inferences
in large scale `omics datasets\textsuperscript{6,18--22}.

@@

Such is not the case for MR, where in order to infer the direction of
causality the instrument must be known to influence the exposure
directly, associating with the outcome only through the exposure. This
requirement can be problematic because if there exists a putative
association between two variables, with the instrumental SNP being
robustly associated with each, it can be difficult to determine which of
the two variables is subject to the direct effect of the SNP (i.e., for
which of the two variables is the SNP a valid instrument? Figure 1). By
definition, we expect that if the association is causal then a SNP for
the exposure will be associated with the outcome, so if the researcher
erroneously uses the SNP as an instrument for the outcome then they are
likely to see an apparently robust causal association of outcome on
exposure. Such a situation can arise in many scenarios. Genome-wide
association studies (GWASs) that identify genetic associations for
complex traits are, by design, hypothesis free and agnostic of genomic
function, and it often takes years of follow up studies to understand
the biological nature of a putative GWAS hit\textsuperscript{23}. For
example, the same FTO variant that is associated with body mass index
(BMI) is also associated with C-reactive protein (CRP). It is known that
this is because BMI causes CRP\textsuperscript{24}, but supposing we
only had \emph{a priori} knowledge that FTO associated with CRP, and we
used it to instrument CRP in an MR analysis of CRP against BMI, we would
mistakenly infer that CRP causes BMI. Another situation is where the
causal directions between 'omic measures need to be determined, for
example if a DNA methylation probe is associated with expression of an
adjacent gene, then is a cis-acting SNP an instrument for the DNA
methylation level, or the gene expression level\textsuperscript{25}
(Figure 1)?

MR does however have some important advantages over mediation based
approaches. First, MR is used as a tool to not just deduce the existance
of causality, but to also estimate the causal effect size. Second, using
mediation requires that the exposure, outcome and instrument variables
are all measured in the same data, whereas recent extensions to MR
circumvent this requirement, allowing causal inference to be drawn when
exposure variables and outcome variables are measured in different
samples\textsuperscript{26}. This has the crucial advantage of improving
statistical power by allowing analysis in much larger sample sizes, and
dramatically expands the breadth of possible phenotypic relationships
that can be evaluated. Third, MR is substantially more robust to there
being measurement error in the exposure and outcome variables. Indeed
instrumental variable (IV) analysis was in part initially introduced as
a correction for measurement error\textsuperscript{27}, whereas it has
been noted that both classic mediation-based
analyses\textsuperscript{14,15,28,29} and mediation-based methods that
use instrumental variables\textsuperscript{30,31} are prone to be
unreliable in its presence. We argue that this is a serious limitation.

Using theory and simulations we show how non-differential measurement
error and measurement imprecision can lead to unreliable causal
inference in the mediation-based CIT method. We then present an
extension to MR that allows researchers to ascertain the causal
direction of an association even when the biology of the instruments are
not fully understood, and also a method to evaluate the sensitivity of
the result to measurement error. This improves the utility of MR in
cases where mediation based methods might have otherwise been used
preferentially. Finally, we apply this new method to infer the direction
of causation between DNA methylation levels and gene expression levels.

@@

The analysis proceeds by simulating variables that have known causal
effects and valid instruments, and then applying measurement error in
different configurations. We then assess the performance of the CIT test
and MR analysis for a) inference of a causal relationship and b)
inferring the correct direction of causality. Included in the MR
analysis is a simple extension that is used to make inference of the
causal direction when it is unknown which of the correlated variables
the instrument is directly acting upon. All analysis was performed using
the R programming language\textsuperscript{32} and code is made
available at \href{}{github url to go here}.

@@

The CIT method\textsuperscript{4} is implemented in the R package
\emph{R/cit}\textsuperscript{17}. The methodology of the CIT is as
follows. Assume an exposure \(x\) is instrumented by a SNP \(g\), and
the exposure \(x\) is causally related to an outcome \(y\). Thus,

@@

where \(\hat{x} = \hat{\alpha}_g + \hat{\beta}_g g\), a causal
relationship is inferred if the null hypothesis that \(\beta_{MR} = 0\)
is rejected. A p-value for this test, \(p_{MR}\), was obtained using the
R package \(R/systemfit\)\textsuperscript{33}.

The direction of the causal association was inferred as follows.
Assuming no horizontal pleiotropy (i.e. \(g\) only has an effect on one
of the variables through the other variable) it is desirable to know
which of the variables, \(x\) or \(y\), it has a direct influence on.
This can be achieved by assessing which of the two variables has the
biggest absolute correlation with \(g\) (Appendix 2). This test can be
formalised by testing for a difference in the correlations \(\rho_{gx}\)
and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
within a population\textsuperscript{34}. It is calculated as

@@

Note that the same correlation test approach can be applied to a
two-sample MR\textsuperscript{26} setting. Two-sample MR refers to the
case where the SNP-exposure association and SNP-outcome association are
calculated in different samples (or from publicly available summary
statistics). Here the Steiger test of two independent correlations can
be applied where.

@@

An advantage of using the Steiger test in the two sample context is that
it can compare correlations in independent samples where sample sizes
are different. Steiger test statistics were calculated using the
\emph{r.test} function in the R package
\emph{R/psych}\textsuperscript{35}.

\subsubsection{Steiger sensitivity
analysis}\label{steiger-sensitivity-analysis}

The Steiger test for testing if \(x \rightarrow y\) is based on
evaluating \(\\rho_{gx} > \\rho_{gy}\). However, \(\\rho_{gx}\) and
\(\\rho_{gy}\) are underestimated if \(x\) and \(y\) are measured
imprecisely. If, for example, \(x\) has lower measurement precision than
\(y\) then we might empirically obtain
\(\\rho_{g,x_o} > \\rho_{g,y_o}\). To evaluate how reliable, \(R\), the
Steiger test is to potential measurement imprecision in \(x\) and \(y\)
we calculate the possible measurement error values which are compatible
with the direction of causality obtained empirically from the Steiger
test. We also calculate the measurement error values which are not
compatible. The ratio of these quantities is an indication of how much
of the measurement error space supports the inferred causal direction. A
ratio \(R=1\) indicates that the Steiger test is highly sensitive to
measurement error, with equal weight of the measurement error parameter
space supporting both directions of causality. Full details are provided
in Appendix 2.

@@

Two sample MR\textsuperscript{26} was performed using the summary
statistics presented in\textsuperscript{25} for each of 4122
associations between DNA methylation and gene expression levels. Because
only summary statistics were available (effect, standard error, effect
allele, sample size, p-values) for the instrumental SNP on the
methylation and gene expression levels, the Steiger test of two
independent correlations was used to infer the direction of causality
for each of the associations. The Wald ratio test was then used to
estimate the causal effect size for the estimated direction for each
association.

@@

This is a simple method to orient the direction of causality in an MR
analysis when the underlying biology of the SNP is not fully understood.
But it can be shown that in the presence of measurement imprecision,
\(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\) (Appendix 2), where \(d\)
is a metric of the degree to which the Steiger test is liable to provide
the wrong estimate (\emph{i.e.} if \(d>0\) then the Steiger test is
likely to be correct about the causal direction). Thus it is important
to note that under certain conditions when measurement imprecision is
present the Steiger test could make erroneous inference about the causal
direction because the value of \(d\) is not restricted to being greater
than 0. However Figure 7 shows that when there is no measurement error
in \(x\), the Steiger test is unlikely to infer the wrong direction of
causality even if there is measurement error in \(y\). It also shows
that in most cases where \(x\) is measured with error, especially when
the causal effect between \(x\) and \(y\) is not very large, the
sensitivity of the Steiger test to measurement error is relatively low.

@@

We used the Steiger test to infer the direction of causality between DNA
methylation levels and gene expression levels. To do this we obtained a
list of 458 gene expression - DNA methylation associations as reported
in Shakhbazov et al\textsuperscript{25}. These were filtered to be
located on the same chromosome, have robost correlations after
correcting for multiple testing, and to share a SNP that had a robust
cis-acting effect on both the DNA methylation probe and the gene
expression probe. Using the reported summary data we used the Steiger
test of two independent correlations to identify which direction of
causality was most likely for each association. We found that the causal
direction commonly goes in both directions (Table 1) but methylation
levels were more likely to influence expression levels than vice versa
(probability of expression causing methylation = 0.0163873 (95\% CI
0.3738727 - 0.4874269) amongst NA associations with
\(p_{Steiger} < 0.05\)).

We performed two sample MR\textsuperscript{26} for each association in
the direction of causality that we estimated to be the correct direction
from the Steiger test. We observed that the sign of the MR estimate was
always in the same direction as the Pearson correlation coefficient
reported by Shakhbazov et al\textsuperscript{25}, and that there was a
moderate correlation between the magnitude of the causal correlation and
the observational Pearson correlation (Table 1).

We also observed that for associations where methylation caused gene
expression the causal effect was more likely to be negative than for the
associations where gene expression caused methylation (OR = 0.5326663
(95\% CI 0.3172203 - 0.8851868)), suggesting that reducing gene
expression levels at a controlling CpG typically leads to increased gene
expression levels, consistent with expectation\textsuperscript{36}.

@@

In the epidemiological literature issues of measurement error in
mediation is relatively well explored\textsuperscript{13}. Our analysis
extends this to related methods such as CIT that are widely used in
predominantly 'omic data. These methods are indeed liable to the same
issues as standard mediation based analysis, and specifically we show
that as measurement error in the exposure variable increases, CIT is
likely have reduced statistical power, and liable to infer the wrong
direction of causality. We also demonstrate that, though seemingly
unintuitive, increasing sample size does not resolve the issue, rather
it leads to more extreme p-values for the model that predicts the wrong
direction of causality.

@@

Mediation based network approaches are very well
established\textsuperscript{31} and have a number of extensions that
make them valuable tools, including for example network construction.
But because they are predicated on the basic underlying principles of
mediation they are liable to suffer from the same issues of measurement
error. Recent advances in MR methodology, for example applying MR to
genetical genomics\textsuperscript{37}, multivariate
MR\textsuperscript{38}, network MR\textsuperscript{39} and mediation
through MR\textsuperscript{40} may offer more robust alternatives for
these more complicated problems.

In our simulations we focused on the simple case of a single instrument
in a single sample setting, and assumed that pleiotropy (the influence
of the instrument on the outcome through a mechanism other than the
exposure) was not present. Recent method developments in
MR\textsuperscript{41} have focused on accounting for the issues that
horizontal pleiotropy can introduce, but how they perform in the
presence of measurement error remains to be explored. An important
advantage that MR confers over most mediation based analysis is that it
can be performed in two samples, which can considerably improve power
and expand the scope of analysis. However, whether there is a
substantive difference in two sample MR versus one sample MR in how
measurement error has an effect is not yet fully understood. We have
also assumed no measurement error in the genetic instrument, which is
not unreasonable given the strict QC protocols that ensure high quality
genotype data is available to most studies. We have restricted the scope
to only exploring non-differential measurement error and avoided the
complications incurred if measurement error in the exposure and outcome
is correlated, however our analysis goes beyond many previous
explorations of measurement error by assessing the impacts of both
imprecision (noise) and linear transformations of the true variable on
causal inference.

@@

\begin{verbatim}
## $r2_exp
## [1] 0.1
## 
## $r2_out
## [1] 0.03162278
## 
## $r2_exp_adj
## [1] 0.1
## 
## $r2_out_adj
## [1] 0.03162278
## 
## $correct_causal_direction
## [1] TRUE
## 
## $steiger_test
## [1] 0
## 
## $correct_causal_direction_adj
## [1] TRUE
## 
## $steiger_test_adj
## [1] 0
## 
## $vz
## [1] -0.08934366
## 
## $vz0
## [1] 0.0262863
## 
## $vz1
## [1] 0.11563
## 
## $sensitivity
## [1] 0.185224
## 
## $sensitivity_ratio
## [1] 4.398868
## 
## $sensitivity_plot
\end{verbatim}

\includegraphics{manuscript_files/figure-latex/steiger_sensitivity_plot-1.pdf}
\includegraphics{manuscript_files/figure-latex/steiger_sensitivity_plot-2.pdf}
\includegraphics{manuscript_files/figure-latex/steiger_sensitivity_plot-3.pdf}

Figure 3: a) We can predict the values the Steiger test would take
(z-axis) for different potential values of measurement error (x and y
axes), drawn here as the blue surface. When \(\rho_{g,y} > \rho_{g,x}\),
as denoted by range of values where the blue surface is above the red
plane, those values of measurement error lead to our observed Steiger
test inferring the wrong causal direction. Where the blue surface lies
below the red plane, these measurement error values support the inferred
causal direction of X to Y. A measure of reliability, therefore, is the
ratio of the negative and positive volumes of the total space bound by
the blue and red surfaces, \(R = \frac{V_{z \geq 0}}{ - V_{z < 0} }\).
In this case, where \(\rho_{g,x}^2 = 0.01\) and \(\rho_{x,y}^2 = 0.1\),
the \(R = 4.40\), which means that 4.40 times as much of the possible
measurement error values are in support of the \(x \rightarrow y\)
direction of causality than \(y \rightarrow x\). b) The influence of
\(\rho_{x,y}\) (x-axis) on the Steiger reliability values (y-axis) for
different values of \(\rho_{g,x}^2\). Inferring the direction of
causality is more sensitive to measurement error when there are stronger
causal associations between X and Y.

@@

In the presence of measurement error in \(x\) and \(y\), however, the
Steiger test will instead be assessing the inequality
\(\rho_{g, x_o} > \rho_{g, y_o}\). In order to assess how reliable the
Steiger test is in the presence of measurement imprecision, we can
evaluate the range of potential values of measurement error in \(x\) and
\(y\) over which the empirical result from the Steiger test would return
the wrong value.

@@

Where \(z = \rho_{g,y} - \rho_{g,x} < 0\) the measurement error range in
\(x_o\) and \(y_o\) will not change the qualitative inference of the
direction of causality. To evaluate the reliability, \(R\), of the
Steiger test to measurement error, the objective is to compare the
weighted proportion of the parameter space that agrees with the inferred
direction against the weighted proportion which does not:

\[
R = \frac{V_{z \geq 0}}{ - V_{z < 0} }
\]

If \(R=1\) then the direction of causality is equally probable across
the range of possible measurement error values. If \(R > 1\) then \(R\)
times as much of the weighted parameter space favours the inferred
direction of causality. \(V_{z}\), the total volume of the function
(Figure 3), can be obtained analytically by solving:

@@

\(V_{z \ge 0}\), the proportion of the volume that lies above the
\(z=0\) plane, can also be obtained analytically. The region of this
volume is bound by the values of \(\rho_{x,x_o}\) and \(\rho_{y,y_o}\)
where \(0 = \rho_{g,y} - \rho_{g,x}\), which can be expanded to
\(\rho_{y,y_o} = \rho_{g,y_o}\rho_{x,x_o} / \rho_{g,x_o}\). Hence,

\[
\begin{aligned}
V_{z \ge 0} & = \int^1_{\rho_{g,x_o}} \int^{\frac{\rho_{g,y_o}\rho_{x,x_o}}{\rho_{g,x_o}}}_{\rho_{g,y_o}} \frac{\rho_{g,y_o}}{\rho_{y,y_o}} - \frac{\rho_{g,x_o}}{\rho_{x,x_o}}\,\,\,\, d\rho_{y,y_o}d\rho_{x,x_o} \\
& = 2\rho_{g,x_o}\rho_{g,y_o} - 2\rho_{g,y_o} - \rho_{g,y_o}log(\rho_{g,x_o}) - \rho_{g,x_o}\rho_{g,y_o}log(\rho_{g,x_o})
\end{aligned}
\]

Thus \(V_{z < 0} = V_{z} - V_{z \geq 0}\).

@@

\hypertarget{refs}{}
\hypertarget{ref-Phillips1991}{}
1. Phillips, A. N. \& Davey Smith, G. How independent are `independent'
effects? relative risk estimation when correlated exposures are measured
imprecisely. \emph{Journal of Clinical Epidemiology} \textbf{44,}
1223--1231 (1991).
